Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2015-09-13 Documents/Paper/ms.tex
portillo
Some notation changes

\newcommand{\hyperparams}{\boldsymbol{\alpha}}
\newcommand{\xx}{\mathbf{x}}

@@

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
components in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\mathcal{D}$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass, highlighting the need for
Nested Sampling.}

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the conditional prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\mathcal{D}$, we usually want to calculate properties of the
posterior distribution, given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \mathcal{D}) \propto
p(N, \hyperparams, \{\xx_i\})
p(\mathcal{D} | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The structure of the model is depicted graphically in Figure~\ref{fig:pgm}.

@@

The marginal likelihood, or evidence, is also an important quantity, and is
the normalisation constant of the posterior given in Equation~\ref{eq:bayes}:
\begin{eqnarray}
\mathcal{Z} &=& p(\mathcal{D})\\
&=& \sum_{i=0}^{N_{\rm max}} \int
p(N, \hyperparams, \{\xx_i\})p(\mathcal{D} | N, \hyperparams, \{\xx_i\})
\, d^N \xx_i \, d\hyperparams \label{eq:evidence}
\end{eqnarray}

@@

\section*{Acknowledgements}
This work is supported by a Marsden Fast-Start grant
from the Royal Society of New Zealand. I would like to thank the following
people for valuable conversations and inspiration:
Anna Pancoast (UCSB), David Hogg (NYU), Daniel Foreman-Mackey (NYU),
Courtney Donovan (Auckland), Tom Loredo (Cornell), Iain Murray (Edinburgh),
Ewan Cameron (Oxford),
John Skilling (MaxEnt Data Consultants), and Daniela Huppenkothen
(Amsterdam, NYU).

>>>>

\newcommand{\hyperparams}{\boldsymbol{\alpha}}
\newcommand{\xx}{\boldsymbol{x}}
\newcommand{\data}{\boldsymbol{D}}

@@

\begin{document}
\maketitle
\abstract{Many inference problems involve inferring the number $N$ of
components in some region, along with their properties $\{\xx_i\}_{i=1}^N$,
from a dataset $\data$. A common statistical example is finite mixture
modelling. In the Bayesian framework, these problems are typically solved using
one of the following two methods: i)
by executing a Monte Carlo algorithm (such as Nested Sampling) once for each
possible value of $N$,
and calculating the marginal likelihood or evidence as a function of $N$; or
ii) by doing a single run that allows the model dimension $N$ to change (such as
Markov Chain Monte Carlo with birth/death moves), and obtaining the posterior
for $N$ directly. In this paper we present a general approach to this problem
that uses trans-dimensional MCMC embedded {\it within} a Nested Sampling
algorithm, allowing us to explore the posterior distribution and calculate
the marginal likelihood (summed over $N$) even if the problem contains a phase
transition or other difficult features such as multimodality.
We present two example problems, finding sinusoidal signals in
noisy data, and finding and measuring galaxies in a noisy astronomical image.
Both of the examples demonstrate phase transitions in the relationship between
the likelihood and the cumulative prior mass, highlighting the need for
Nested Sampling.}

@@

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the conditional prior for $\{\xx_i\}$ is iid and does not depend on $N$.
Given data $\data$, we usually want to calculate properties of the
posterior distribution, given by:
\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\} | \data) \propto
p(N, \hyperparams, \{\xx_i\})
p(\data | N, \hyperparams, \{\xx_i\})\label{eq:bayes}
\end{eqnarray}
The structure of the model is depicted graphically in Figure~\ref{fig:pgm}.

@@

The marginal likelihood, or evidence, is also an important quantity, and is
the normalisation constant of the posterior given in Equation~\ref{eq:bayes}:
\begin{eqnarray}
\mathcal{Z} &=& p(\data)\\
&=& \sum_{i=0}^{N_{\rm max}} \int
p(N, \hyperparams, \{\xx_i\})p(\data | N, \hyperparams, \{\xx_i\})
\, d^N \xx_i \, d\hyperparams \label{eq:evidence}
\end{eqnarray}

@@

\section*{Acknowledgements}
This work is supported by a Marsden Fast-Start grant
from the Royal Society of New Zealand. I would like to thank the following
people for valuable conversations and inspiration:
Anna Pancoast (UCSB), David Hogg (NYU), Daniel Foreman-Mackey (NYU),
Courtney Donovan (Auckland), Tom Loredo (Cornell), Iain Murray (Edinburgh),
Ewan Cameron (Oxford),
John Skilling (MaxEnt Data Consultants), and Daniela Huppenkothen
(Amsterdam, NYU). Stephen Portillo (Harvard) discovered a significant bug.
